CRUSE to Safe Cycling in Ireland \\\large{An Open Source Tool to Support Active Travel}

### 2024-03-25 paper-v1.tex

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{European Transport Research Review}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Cycle Route Uptake and Scenario Estimation (CRUSE): an approach for developing strategic cycle network planning tools},
  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Toole; Dan Brennan; Suzanne Meade},
  pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Cycle Route Uptake and Scenario Estimation (CRUSE): an approach
for developing strategic cycle network planning tools}
\author[1]{Robin Lovelace%
\corref{cor1}%
}
 \ead{r.lovelace@leeds.ac.uk} 
\author[1]{Joey Talbot%
%
}

@@

\author[2]{Gary O'Toole%
%
}

@@

\begin{abstract}
This paper describes an approach for developing strategic cycle network
planning tools. Based on our experience developing and deploying the
Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland, we
outline the underlying methods, including disaggregation of
origin-destination data, incorporation of additional trip purposes,
routing, scenario generation, and development of an intuitive user
interface. Commissioned and used by Transport Infrastructure Ireland,
CRUSE provides estimates of current and potential future cycling levels
under `snapshot' scenarios to inform investment decisions. The publicly
available results at https://cruse.bike/ enable planners, engineers, and
other stakeholders to make more evidence-based decisions. CRUSE goes
beyond previous work by: modeling networks at high spatial resolution;
simulating multiple trip purposes (social, shopping, personal utility,
recreational, and cycle touring) in addition to using official
origin-destination data; and providing estimates of `quietness' (a proxy
for cyclist comfort and route preference) at the route segment level.
Three network types --- `Fastest', `Balanced', and `Quietest' --- help
plan both arterial and residential cycle networks. Workshops and
unstructured interviews with stakeholders show that the tool is already
being used in practice to inform urban, inter-urban, and rural network
planning. The approach is flexible and open-source, allowing the
underlying ideas and code to be adapted, supporting effective cycling
policies and interventions internationally.
\end{abstract}

@@

Transport systems dominated by heavy and powerful private cars are
inefficient, dangerous, and unhealthy. Cars are a major and growing
source of greenhouse gas emissions causing climate change
\citep{winkler2023}, a leading cause of premature death and injury due
to road traffic collisions \citep{globals2018} , and a cause of disease
due to physical inactivity and air, noise and (from tyres) microplastic
pollution \citep{mattsson2023} \citep{welch2023} \citep{cavallaro2024}.
Transport is responsible for 23\% of global greenhouse gas (GHG)
emissions, 70\% of which are from road transport, with passenger cars
accounting for nearly half of transport emissions (around 10\% of global
emissions) \citep{jaramillo2022}. The transport system encourages,
enables and in some cases enforces unsustainable lifestyles. Services
that are only accessible by car lock-in car dependency
\citep{gray2001, shergold2012, motte-baumvol2010}.

Growing evidence of the negative impacts of car-dependent transport
systems has led governments in many countries to set targets and take
actions. In the context of climate, road safety and physicial inactivity
crises, policies to improve transport systems can be classified
according to the `Avoid-Shift-Improve' (ASI) framework
\citep{jaramillo2022}. The ASI framework highlights the importance of
demand reduction (\emph{avoid}ing unnecessary trips), in addition to
mode \emph{shift} to sustainable modes and \emph{improvement} of
existing energy converters, in that order.

@@

Another motivation for cycle network planning at the European level is
the Road Infrastructure Safety Management (RISM) directive (2008/96/EC),
which requires member states to implement a road safety management
system (RSMS) for all public roads. Specifically, ``Member States shall
ensure that the ranking of high accident concentration sections and the
network safety ranking are carried out'' \citep{directiv2008}. Given
that `safety' in this context is usefully quantified as the number of
people killed or seriously injured (KSI) per distance traveled, the
directive requires estimation of distance traveled by mode, down to the
road link level. For active modes, about which there is a paucity of
data compared with motorized modes, this is a major challenge. Better
data to inform road safety policies and interventions is a motivation
for better estimates of `baseline' levels of physical activity at high
geographic resolutions \citep{tait2023}.

National governments are increasingly acting on the evidence. In
Ireland, the Road Safety Authority (RSA) has set the target of halving
the number of road traffic deaths and serious injuries by 2030
\citep{national2021}. Doing so while simultaneously enabling rapid
uptake of active modes will require key travel corridors to be
identified and `cycle proofed'. Cycling in Ireland represents only 3\%
of total modal share as of the 2016 Census, but accounts for 20\% of
serious injuries and 7\% of all fatalities. Poor perception of safety
has been found to represent the most important barrier to increased
cycling in Ireland\citep{brick2018}, and the need to improve road safety
drives the development of national cycling policy and infrastructure
plans. The main frameworks underpinning these efforts are the Climate
Action Plan \citep{climate2022a}, the National Development Plan, and the
National Roads 2040 strategy \citep{national2023}.

At the regional level within Ireland, the recently published Greater
Dublin Area Transport Strategy \citep{greater} reveals the high support
for and capability for cycling: ``nearly a quarter of adults cycle at
least once a week in the Dublin Metropolitan Area'' with cycling in the
Dublin area taking up to 60,000 cars off the road today. Extrapolating
this on a per population basis across Ireland, with around 40\% of the
population living in Dublin, suggests that around 150,000 cars could be
removed nationwide just by achieving Dublin levels of cycling in all
counties (notwithstanding existing cycling trips and differences in trip
distances). 7 in 10 trips in Ireland are by car \citep{national} .
Cycling has the potential to replace a large proportion of these trips:
``A high priority must also be given to cyclists, because trips by this
mode have the potential to replace trips by private car, most
specifically for short to medium distance trips, but increasingly for
longer trips as ebikes extend the range of this mode'' \citep{greater}.

Further evidence of the importance of cycling in Ireland is provided by
the National Strategic Objective (NSO) from the National Development
Plan, which allocates €8.6 billion to sustainable transport
infrastructure including public transport and active travel
interventions. Cycle infrastructure will be developed in synchrony with
the BusConnects project, an entire redesign of the bus network in Dublin
and Cork. It was in this context that Transport Infrastructure Ireland
(TII) commissioned the research reported in this paper, which led to the
Cycle Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland.
Building on previous work, including the Propensity to Cycle Tool (PCT)
for England and Wales, the CRUSE Tool was developed to provide evidence
on current cycling levels and future cycling potential nationwide across
Ireland.

Stakeholder consultation by TII emphasized the need for a strong,
national, systematic but locally-specific evidence base on cycle
networks in both urban and rural Ireland. The evidence base had to scale
nationally to support strategic alignment with national, regional and
local policies, but also be useful for local network planning.

@@

\includegraphics{images/paste-10.png}

@@

\caption{\label{fig-landing}Illustration of the landing page of the
CRUSE Tool for Ireland.}

@@

This paper describes an approach to strategic cycle network planning
tool development that is open, scalable, and evidence-based. We label
the approach Cycle Route Uptake and Scenario Estimation (CRUSE) and
present a case study of its application in Ireland, the results of which
are publicly available at \href{https://cruse.bike/}{cruse.bike} in the
CRUSE Tool for Ireland, the landing page of which is illustrated in
Figure~\ref{fig-landing}. In Section~\ref{sec-methods}, we outline the
methods used to generate the evidence presented in the CRUSE Tool. In
Section~\ref{sec-results}, we present the results of the CRUSE Tool,
including estimates of current cycling levels and future cycling
potential at the national, regional and local levels. In
Section~\ref{sec-discussion}, we discuss limitations and possible future
improvements to the approach, and the implications of the results for
Ireland before making concluding remarks in
Section~\ref{sec-conclusions}.

\section{Tools for estimating cycling
potential}\label{tools-for-estimating-cycling-potential}

Tools for estimating geographical distribution of cycling potential have
advanced substantially in recent years. Developed by researchers and
developers in academic, public and private sectors, they have evolved
from simple area-based static models tailored to specific regions to
more complex and (in some cases) more generalizable tools. These tools
can be considered as a specific application of the concept of planning
support systems (PSS) , which are designed to help planners and
decision-makers to make more informed decisions \citep{geertman2009}.
Outputs include maps with results available at several levels of
analysis with levels of availability ranging from being only available
to researchers and in static maps to open access mapping systems
available to the public.

Previous studies have applied the concept of PSS to specific modes,
including walking \citep{bencekri2024}, public transport
\citep{barmentlo2012} and cycling \citep{bencekri2023}. While such prior
work provides insight into the potential for approaches to transport
planning that are both data-driven and participatory, the focus of this
paper is on tools that focus attention on the \emph{geographic
distribution} of cycling potential and which answer the question ``where
to build''. Not all papers reviewed in this section classify themselves
as, or even mention, PSS, but all of them can be considered as PSS. This
paper is informed by the thinking underlying PSS, including the idea
that the resulting tools are more effective if they are modular and can
inter-link with other tools and processes in the planning process: PSS
can be conceived as a ``toolbox with separate instruments that can work
and talk to each other effectively (as in the open systems concept) and
that will be selectively applied in varying configurations to support a
particular phase in the planning process'' \citep{geertman2002}. We
share this conception of PSS, emphasizing the benefits of a focus on a
particular mode (cycling) and a particular phase in the planning process
(strategic network development).

Some of the most relevant existing tools are summarised below and
presented in Table~\ref{tbl-tools}.

\begin{longtable}[t]{>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{15em}>{\raggedright\arraybackslash}p{12em}>{\raggedright\arraybackslash}p{5em}}

\caption{\label{tbl-tools}Summary of existing tools for estimating
cycling potential.}

\tabularnewline

\\
\toprule
Approach & Coverage and input data & Outputs and availability & License\\
\midrule
\cellcolor{gray!10}{Bicycle share model, Parkin et al. (2007)} & \cellcolor{gray!10}{England and Wales, Journeys to work OD census data, at the small-area (wards) level} & \cellcolor{gray!10}{Static tables in academic paper} & \cellcolor{gray!10}{Proprietary}\\
Analysis of  Cycling Potential (ACP), Transport for London (2010, 2016) & London, London Travel Demand Survey & Static tables, graphs, and heatmaps, not publicly available & Proprietary\\
\cellcolor{gray!10}{Prioritization index, Larsen et al. (2013)} & \cellcolor{gray!10}{Montreal, Survey, Road safety, and OD data} & \cellcolor{gray!10}{Map-based heatmap, not publicly available} & \cellcolor{gray!10}{Proprietary}\\
Usage intensity index, Zhang et al. (2014) & Belo Horizonte, Survey, census, and OD data & Static tables and maps, not publicly available & Proprietary\\
\cellcolor{gray!10}{The Cycling Potential Tool (CPT), Phillips and Range (2017)} & \cellcolor{gray!10}{Scotland, Environmental and socioeconomic data, at the small area (output areas) level} & \cellcolor{gray!10}{Maps showing cycling potential in each area} & \cellcolor{gray!10}{Proprietary}\\
\addlinespace
Propensity to Cycle Tool (PCT), Lovelace et al. (2017) & England and Wales, Journeys to work and school OD census data, at the small area (LSOA) level & Online maps, graphs, tables, publicly available at www.pct.bike & Open source\\
\cellcolor{gray!10}{The    Gross Potential for Cycling tool (CPC), Silva et al. (2021 and 2022)} & \cellcolor{gray!10}{21 Portuguese cities, Land use and socio-demographic data, at the small area (census tract) level} & \cellcolor{gray!10}{Static maps showing cycling potential in different areas} & \cellcolor{gray!10}{NA}\\
\bottomrule

\end{longtable}

Area-based approaches were developed before network-based tools, such as
a 2007 regression-based model to estimate the percentage of cycling
trips to work in England and Wales based on socioeconomic,
transportation, and physical factors at the small area (wards) level in
2007 \citep{parkin_estimation_2007}. They found that areas with a higher
percentage of females, non-whites, car ownership, lower socioeconomic
classes, income, distance to work, population density, poor highway
conditions, hills, rainfall, and fewer off-road bicycle routes tended to
have a lower proportion of cycling to work. The approach also generated
``forecasts for potential levels of bicycle use'' with with increases in
off-road routes, improvements of highway and, conversely, increases in
car ownership and distance to work leading to decreases in cycling to
work.

Three years later, Transport for London created the Analysis of Cycling
Potential (ACP) tool based on a mode choice model that assesses the
likelihood of trips being cycled, with data from the TfL London Travel
Demand Survey as its primary input. The tool generates tables, graphs,
and heatmaps showing the potential for cycling for different trips
purposes in London to guide local cycling initiatives, such as where new
hire `docking stations' should go. The tool does not use
origin-destination data and cannot estimate cycling potential on
specific routes \citep{transport_for_london_analysis_2010}.

Other tools that generate areal data to prioritize investment include a
study that generated a `prioritization index' for Montreal, Canada
\citep{larsen_build_2013}, a `usage intensity index' based on stated
preferences in Belo Horizonte, Brazil, and a Cycling Potential Tool
(CPT) for Scotland \citep{phillips_development_2017}. The CPT is
composed of the base environmental module --- based on eight weighted
factors (population density, hilliness, physical barriers, access to
services, existing cycling mode share, distance to work and school, and
road speed) --- and the quality of service module, which evaluates each
area of interest based on eight factors (surface condition, adjacent
cyclists, comfort factor, conflict, distance between junctions, slope,
access to services, and origin/destination).

A more recent areal-based approach is the designed the Gross Potential
for Cycling (GPC) tool to prioritize areas for cycling infrastructure
and other cycling measures \citep{silva_gross_2021}. The GPC uses two
sets of indicators: population-based indicators (age, potential demand
density, employment density and motorization rate) and area-based
indicators (accessibility, education, public transport, connectivity,
land use mix and relative performance). These indicators are calculated
considering topography, road hierarchy, and average congestion and
presented on a scale from 5 to 10. They are then combined into an
overall score, weighted by their impact on cycling according to the
literature. The ranking of areas by the GPC are
\href{https://boost.up.pt/en/ferramentas/gpc}{publicly available} and
its practical value has been assessed through workshops
\citep{silva_revealing_2022}.

Route-based tools emerged with increasing availability of
origin-destination data, routing engines that can assign trips to
networks, and improvements in computer hardware and software needed to
generate route networks. A prominent example is the the Propensity to
Cycle Tool (PCT), first developed to estimate current and future levels
of cycling at desire line, zone, route, and route network levels for
case study cities \citep{lovelace_mapping_2016}. The approach was
scaled-up to estimate estimate the potential benefits of uptake at zone
and desire line levels nationally, and launched as a publicly available
web application in 2017 \citet{lovelace2017}. Extensions of the PCT
approach have included estimation of benefits at the individual level
\citep{woodcock2018}, addition of travel to school network
\citep{goodman2019}, and improved modeling of impacts on health,
environmental and distributional outcomes \citep{woodcock2021}.
Initially developed just for England, the PCT was extended to cover all
of Wales (for commuter data only) in 2018.

@@

The approach presented in this paper seeks to overcome three limitations
of previous tools to estimate cycling potential: 1) low resolution of
data, with routes starting and ending in administrative zone centroids,
2) limited coverage of trip purposes beyond travel to work and school,
and 3) a web interface that was not user-friendly or intuitive.

@@

\includegraphics{images/paste-8.png}
\includegraphics{images/paste-9.png}

\caption{\label{fig-dublin}Illustration of the density of route networks
generated by the CRUSE Tool for Dublin City and surroundings (top) and
rural County Mayo on the West coast of Ireland (botton). Source: CRUSE
Tool, publicly available at \href{https://cruse.bike/}{cruse.bike}.}

@@

The Climate Action Plan scenario models the transport emissions
reductions targeted in the Irish Government's
\href{https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/}{Climate
Action Plan 2021}, which aims for a 51\% cut in overall greenhouse gas
emissions by 2030 as part of the pathway to net-zero emissions by 2050.
For transport, this includes 500,000 extra walking, cycling and public
transport trips per day by 2030. In terms of car travel, the target is
to ``Increas{[}e{]} the proportion of kilometers driven by passenger
electric cars to between 40 and 45\% by 2030, in addition to a reduction
of 10\% in kilometers driven by the remaining internal combustion engine
cars.'' This equates to a 5.5 to 6\% reduction in total car km driven.

@@

The main result of the work presented in this paper is an open access
web application and datasets on current and potential future cycling
levels in Ireland, with evidence provided at county and route segment
levels. Like other national tools presented in Table~\ref{tbl-tools}, a
key feature of the results is that it provides a consistent and
objective baseline for comparing cycling potentials in different areas.

The approach generates estimates of cycling potential on every road in
the country and are therefore too extensive to present in their entirety
in this paper. Through the interactive web application, users can
explore the data to generate the results that are most relevant to their
needs, whether that is finding the cycling potential on a particular
road or finding `weak links' or barriers in the cycle network associated
with a particular school, work place or other destination.

Instead of trying to show such use cases, of which there are many
hundreds, we present a selection of results to illustrate the main
features of the resulting evidence, in Section~\ref{sec-quantiative}. We
also present some qualitative results based on workshops and interviews
with stakeholders in Ireland, in Section~\ref{sec-qualitative}.

\subsection{Selected quantitative results}\label{sec-quantiative}

@@

\subsection{Selected qualitative results}\label{sec-qualitative}

Thoughout the project, we undertook a series of workshops and interviews
with stakeholders in Ireland to gather feedback on the approach and
resulting tool. The feedback was generally positive, with stakeholders
commenting on the need for the tool, lack of data on cycling potential
proving a barrier to decision-making and investment, and the potential
for the tool. At an end of project workshop in December 2023,
stakeholders were asked to provide feedback on the tool and its
potential use in practice. Quotes and suggestions from these workshops,
and a subsequent interview with a Senior Transport Planning using the
evidence in their everyday work in Transport Infrastructure Ireland, and
presented in this section (with quotes being attributed to named
individuals where permission has been granted).

Early in the project, we presented the approach to practitioners in
Kildare and Limmerick. The feedback from the workshop in Kildare
highlighted the need for the tool for practitioners at the county level,
as highlighted in the following quote:

\begin{quote}
``It's a missing piece of evidence that will help new projects get
built'' (Senior Engineer, Sustainable Transport \& Traffic Management,
Kildare County Council)
\end{quote}

Practitioners in Kildare County Council that the evidence generated
\emph{was already valuable} the development of new cycle routes,
especially when making the case during planning applications. They
emphasized that cycle networks are often fragmented, and the tool allows
them to demonstrate how individual local schemes can contribute to a
larger, interconnected network. Council officers expressed a preference
for visualizations that directly compare the fast and quiet routes, as
they recognize the importance of both types of routes (users can switch
between the on the `Route types' page for each county although this
functionality is not available in the landing page map due to other
feedback requesting a simpler interface).

In comparison with another tool (the National Transport Authority's
cycling propensity tool), the CRUSE tool was seen by practitioners in
Kildare as more user-friendly and intuitive, with the ability to
generate evidence at the route level, which is crucial for planning new
cycling infrastructure. Overall, the workshop participants concluded
that the approach filled gap in evidence and will greatly assist in the
development of new cycling projects.

During the workshop with practitioners in Limerick, the estimates of
Baseline and Go Dutch potential at the network levels were found to
match well with local knowledge. The tool effectively highlighted the
same sections on the transport network that experienced planners
identified as priorities for investment in cycling, such as the South
Circular Road. Limerick County Council expressed their interest in
utilizing each network layer, particularly the quiet network, to support
new cyclists who lack confidence to share space with motor traffic.
Specific suggestions from the workshop included the suggestion to align
scenarios with national strategy (resulting in the Climate Action Plan
scenario) and to provide data downloads provided as `Shapefiles' for
further analysis (now implemented although the download format is
GeoPackage, an open standard for spatial data).

>>>>

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{European Transport Research Review}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={CRUSE to Safe Cycling in Ireland},
  pdfauthor={Robin Lovelace; Joey Talbot; Eugeni Vidal-Tortosa; Hussein Mahfouz; Elaine Brick; Peter Wright; Gary O'Tool; Dan Brennan; Suzanne Meade},
  pdfkeywords={Cycling, Open-source, Road Safety, Active Travel},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{CRUSE to Safe Cycling in Ireland \\\large{An Open Source Tool to
Support Active Travel} }
\author[1]{Robin Lovelace%
\corref{cor1}%
}
 \ead{r.lovelace@leeds.ac.uk} 
\author[1]{Joey Talbot%
%
}

@@

\author[2]{Gary O'Tool%
%
}

@@

\begin{abstract}
Legislation mandates improved road safety and decarbonisation of
predominantly road-based transport network in Ireland and other
countries. Increasing cycling transport, making it safer and more
attractive, is essential to this end. Cycling in Ireland represents only
3\% of total modal share as of the latest Census (2016), but accounts
for 20\% of serious injuries and 7\% of all fatalities. A planning tool
based on high resolution datasets on baseline and potential cycling
levels could greatly assist in achieving this goal.

This paper describes the design, features and potential use of the Cycle
Route Uptake and Scenario Estimation (CRUSE) Tool for Ireland was
commissioned to support cycle network planning, safety interventions,
business cases and project evaluation. The resulting web application is
available at \url{https://cruse.bike/}, enabling planners, engineers,
and other stakeholders to make more evidence-based decisions. CRUSE goes
beyond previous work in several important ways. It models networks at
high levels of geographic resolution, simulates a wide range of trips
including recreational, social, shopping and personal trips, in addition
to trips to work and primary, secondary and tertiary education
locations. The tool provides key `denominator' information to enable
reporting of collision rates per billion km cycled.

Another key feature of CRUSE is its use of frequently-updated
OpenStreetMap data for route calculation and ``cycle friendliness''
estimates of all links on the network. CRUSE provides three network
types: the `Fastest' network highlights routes for directness;
`Quietest' prioritises cycle friendliness, and `Balanced' represents a
balance between speed and comfort. Trips are generated based on origin
and destination data from the 2016 Census in combination with modeled
demand data to estimate current and potential future cycling levels at
area, route, and network levels for each county in Ireland.

For investment in cycling to deliver on the huge potential benefits,
cycling must become safer. With growth in the ebike market, CRUSE will
help inform inter-urban and rural networks to support the transfer of
trips to sustainable modes for longer journeys. Provision of open access
data in a publicly available web tool will address the challenges faced
by counties across Ireland in meeting their sustainability commitments.
The experience of developing a deploying CRUSE are relevant to countries
with sustainable transport policies worldwide.
\end{abstract}

@@

High energy transport systems are a major contributor to climate change,
a leading cause of premature death and injury due to road traffic
collisions, and a cause of disease due to airborne and noise pollutants.
Conversely, evidence-based and effective transport \emph{policies} have
great potential to decarbonize economies, improve public health, and
save lives.

Transport is responsible for 23\% of global greenhouse gas (GHG)
emissions, 70\% of which are from road transport. Nearly half of these
emissions (around 10\% of global emissions) are from passenger cars
\citep{jaramillo2022}. The transport system encourages, enables and in
some cases enforces unsustainable lifestyles. Services that are only
accessible by car lock-in car dependency
\citep{gray2001, shergold2012, motte-baumvol2010}.

Recognizing the growing evidence of such impacts of poorly designed and
performing transport systems, governments in many countries have set
targets and taken actions. In the context of climate, road safety and
physicial inactivity crises, policies to improve transport systems can
be classified according to the `Avoid-Shift-Improve' (ASI) framework
\citep{jaramillo2022}. The ASI framework highlights the importance of
demand reduction (\emph{avoid}ing unnecessary trips), in addition to
mode \emph{shift} to sustainable modes and \emph{improvement} of
existing energy converters, in that order.

@@

Another motivation for cycle network planning is the Road Infrastructure
Safety Management (RISM) directive (2008/96/EC), which requires member
states to implement a road safety management system (RSMS) for all
public roads. Specifically, ``Member States shall ensure that the
ranking of high accident concentration sections and the network safety
ranking are carried out'' \citep{directiv2008}. Given that `safety' in
this context is usefully quantified as the number of people killed and
seriously injured (KSI) per distance traveled, the directive requires
estimation of distance traveled by mode, down to the road link level.
For active modes, about which there is a paucity of data compared with
motorized modes, this is a major challenge. Better data to inform road
safety policies and interventions is a motivation for the CRUSE Tool
outlined in this paper, and modeling active travel more broadly.

In Ireland, the Road Safety Authority (RSA) has set the target of
halving the number of road traffic deaths and serious injuries by 2030
\citep{national2021}. Doing so while simultaneously enabling rapid
uptake of active modes will require key active travel routes to be
identified and improved. A proactive approach to the development of
cycling policy and infrastructure is taking place in Ireland. The main
frameworks underpinning these efforts are the Climate Action Plan
\citep{climate2022a}, the National Development Plan, and the National
Roads 2040 strategy \citep{national2023}. These policy frameworks are
backed-up by public support for active travel interventions: while only
3\% of trips in Ireland are made by cycling according to the latest
Census data from 2016 ``83\% of respondents supported enhanced walking
and cycling facilities'' \citep{sustaina2019}.

At the regional level, the recently published Greater Dublin Area
Transport Strategy \citep{greater} reinforces these findings: ``nearly a
quarter of adults cycle at least once a week in the Dublin Metropolitan
Area'' with cycling in the Dublin area taking up to 60,000 cars off the
road today. Extrapolating this on a per population basis across Ireland,
with around 40\% of the population living in Dublin, suggests that
around 150,000 cars could be removed nationwide just by achieving Dublin
levels of cycling in all counties (notwithstanding existing cycling
trips and differences in trip distances). 7 in 10 trips in Ireland are
by car \citep{national} . Cycling has the potential to replace a large
proportion of these trips: ``A high priority must also be given to
cyclists, because trips by this mode have the potential to replace trips
by private car, most specifically for short to medium distance trips,
but increasingly for longer trips as ebikes extend the range of this
mode'' \citep{greater}.

Further evidence of the importance of cycling in Ireland \emph{already}
is provided by the National Strategic Objective (NSO) from the National
Development Plan, with €8.6 billion allocated to sustainable transport
infrastructure including public transport and active travel
interventions. Cycle infrastructure will be developed in synchrony with
the BusConnects project, an entire redesign of the bus network in Dublin
and Cork.

It was in this context that Transport Infrastructure Ireland (TII)
commissioned the Cycle Route Uptake and Scenario Estimation (CRUSE) Tool
for Ireland. Based on the Propensity to Cycle Tool (PCT) for England and
Wales, the CRUSE Tool was developed to provide evidence on current
cycling levels and future cycling potential nationwide across Ireland.

A key consideration for TII is to ensure both urban and rural Ireland
were targeted for cycling infrastructure, as shown in
Figure~\ref{fig-dublin}. To this end, TII extensively reviewed activity
on Irish Roads and how the road network was being used by commuters,
cyclists, heavy goods vehicles and other road users across the Irish
transport network. The aim was for the tool to provide strong, national,
systematic but locally-specific evidence to monitor cycling friendliness
and safety in Ireland, to ensure strategic alignment with national,
regional and local policies. The open source and publicly available
nature of the tool ensures more inclusive and evidence-based
conversations around cycle network planning between all stakeholders in
the transport planning process.

TII also requires evidence to support monitoring of growth in cycling,
upgrades of existing cycling infrastructure, business case development
and appraisal, and exposure information to estimate crash rates.
Exposure information is a key part of the the European RISM Directive,
for reporting collision rates of vulnerable road users by 2024. These
factors mean that the tool can be seen as an open access `leverage
point', providing key information for and supporting ambitious plans in
many aspects the planning system, from network design to post-build
monitoring \citep{lovelace2020}.

@@

\includegraphics{images/paste-1.png}

\includegraphics{images/paste-3.png}

@@

\caption{\label{fig-dublin}Illustration of the density of route networks
generated by the CRUSE Tool for Dublin City and surroundings (top) and
rural County Mayo on the West coast of Ireland. Source: CRUSE Tool,
available at \url{https://cruse.bike/}.}

@@

This paper aims to describe the design, features and potential use of
the CRUSE Tool. In Section~\ref{sec-methods}, we outline the methods
used to generate the evidence presented in the CRUSE Tool. In
Section~\ref{sec-results}, we present the results of the CRUSE Tool,
including estimates of current cycling levels and future cycling
potential at the national, regional and local levels. In
Section~\ref{sec-discussion}, we discuss the implications of the results
for policy and practice. In Section~\ref{sec-conclusions}, we conclude
the paper.

The methods used to generate the evidence presented in the CRUSE Tool
build on the Propensity to Cycle Tool (PCT), which was originally funded
by the UK's Department for Transport and developed by a multi-university
team. An important feature of the PCT is that it is open source and
publicly available (at \href{https://www.pct.bike/}{www.pct.bike}),
allowing its use by all stakeholders in the transport planning process
\citep{lovelace2017}. The PCT approach has had major policy impacts, as
outlined in Research Excellence Framework (REF) impact case studies,
which demonstrate that the tool ``revolutionized strategic cycle
planning in England and Wales'' by overcoming the barriers to cycling
investment imposed by lack of evidence on cycling
potential\citep{lovelace2023}.

The first version of the PCT was based on current and future potential
uptake of cycling for single stage \emph{travel to work} at desire line,
zone, route, and route network levels \citep{lovelace2016}. It was
launched in April 2017 as the government-endorsed tool for strategic
cycle network planning and part of the Cycling and Walking Investment
Strategy \citep{cycling2017}. Extensions of the PCT approach have
included estimation of benefits at the individual level
\citep{woodcock2018}, addition of travel to school network
\citep{goodman2019}, and improved modeling of impacts on health,
environmental and distributional outcomes \citep{woodcock2021}.
Initially developed just for England, the PCT was extended to cover all
of Wales (for commuter data only) in 2018.

@@

The CRUSE Tool seeks to overcome the following methodological
limitations of the original PCT:

\begin{itemize}
\tightlist
\item
  Low resolution of data, with routes starting and ending in
  administrative zone centroids
\item
  Limited coverage of trip purposes beyond travel to work and school
\item
  A web interface that was not user-friendly or intuitive
\end{itemize}

Methods were developed to overcome each of these, as outlined in
Section~\ref{sec-trip-purposes} to Section~\ref{sec-ui}.

@@

@@

The Climate Action Plan scenario is loosely based on the Irish
Government's
\href{https://www.gov.ie/en/publication/6223e-climate-action-plan-2021/}{Climate
Action Plan 2021}, which contains policies for action to achieve a 51\%
reduction in overall GHG emissions by 2030, on a path towards net-zero
by 2050. For transport, this includes 500,000 extra walking, cycling and
public transport trips per day by 2030. In terms of car travel, the
target is to ``Increas{[}e{]} the proportion of kilometers driven by
passenger electric cars to between 40 and 45\% by 2030, in addition to a
reduction of 10\% in kilometers driven by the remaining internal
combustion engine cars.'' This equates to a 5.5 to 6\% reduction in
total car km driven.

@@

The main result of the work presented in this paper is an open access
web application and datasets on current and potential future cycling
levels in Ireland, with evidence provided at county and route segment
levels. Like the PCT on which CRUSE builds, a key feature of the results
is that they are systematic and available for all counties, and most
routes on which cycling is permitted, in Ireland. The full set of
results are therefore too extensive to present in their entirety in this
paper. Through the interactive web application, users can explore the
data to generate the results that are most relevant to their needs,
whether that is finding the cycling potential on a particular road or
finding `weak links' or barriers in the cycle network associated with a
particular school, work place or other destination.

Instead of trying to show such use cases, of which there are many
hundreds, we present a selection of results to illustrate the main
features of the CRUSE Tool.

@@
